---
title: "objectExplorerR: Interactive Ontology Object Explorer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{objectExplorerR: Interactive Ontology Object Explorer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`objectExplorerR` is the interactive user-facing application in the ontology R
ecosystem. It provides a Shiny-based exploratory UI that lets users search,
filter, traverse links, and inspect ontology objects visually — without writing
code.

This corresponds to Palantir's Object Explorer concept: visual query composition
and exploration views with relationship navigation.

## Quick Start

```{r launch-example, eval=FALSE}
library(ontologySpecR)
library(objectSetsR)
library(actionTypesR)
library(objectExplorerR)
library(DBI)

# Load an ontology bundle
b <- read_bundle("path/to/your-bundle.json")

# Connect to your database
con <- dbConnect(RSQLite::SQLite(), "your-database.db")

# Create ActionContext for action execution (recommended)
action_ctx <- action_context(b, con)
action_ctx <- register_handler(action_ctx, "MyAction", function(...) { ... })

# Launch the explorer
explore_ontology(b, con, action_ctx = action_ctx)
```

## Setup with the Aviation Demo

The ontologySpecR package ships with an aviation demo bundle:

```{r setup-demo, eval=TRUE}
library(ontologySpecR)

# Read the aviation demo bundle
b <- read_bundle(system.file("examples", "aviation-demo.json",
                              package = "ontologySpecR"))

# The bundle defines 3 object types
cat("Object types:", paste(vapply(b$objects, function(o) o$id, ""), collapse = ", "), "\n")

# And 3 link types
cat("Link types:", paste(vapply(b$links, function(l) l$id, ""), collapse = ", "), "\n")

# And 1 action type
cat("Action types:", paste(vapply(b$actions, function(a) a$id, ""), collapse = ", "), "\n")
```

## Application Layout

The explorer is organized into four main panels:

```
+------------------------------------------------------+
|  [Object Type Selector]  [Search: ________]  [☾/☀]  |
+------------------------+-----------------------------+
|                        |                             |
|   FILTER PANEL         |    RESULTS TABLE            |
|                        |                             |
|  Property filters      |    Interactive DT table     |
|  (auto-generated       |    with sort, page,         |
|   from object type     |    select rows              |
|   properties)          |                             |
|                        +-----------------------------+
|  [Apply] [Clear]       |                             |
|                        |    DETAIL PANEL              |
|   LINK PANEL           |    (shown when row selected)|
|                        |                             |
|  Available links       |    - Properties tab         |
|  [Traverse ->]         |    - Links tab              |
|  [<- Search Around]    |    - Actions tab            |
|                        |    - Graph tab              |
+------------------------+-----------------------------+
```

### Top Bar: Object Type Selector + Search + Dark Mode

The top bar contains a dropdown populated from the bundle's object types
(using display names), a free-text search box that filters across all
string properties, and a dark mode toggle button.

### Filter Panel (Left Sidebar)

Filters are **auto-generated** from the selected object type's property
definitions:

| Property Type | Widget                          |
|--------------|---------------------------------|
| `string`     | Multi-select dropdown (≤20 values) or text input |
| `integer`    | Range slider                    |
| `number`     | Range slider                    |
| `boolean`    | Tri-state radio (Any/True/False) |
| `date`       | Date range picker               |
| `datetime`   | Date range picker               |
| `json`       | (not filterable)                |

For string properties with few distinct values (< 20), the filter
automatically uses a multi-select dropdown instead of a text input.

Numeric sliders are initialized with the actual min/max from the data.

### Results Table (Main Area)

An interactive DataTable showing filtered objects with sortable columns,
pagination, and row selection. Clicking a row opens the detail panel.

When concept columns are provided (via `concept_cols_fn`), they appear as
colored badges (green for TRUE, red for FALSE).

### Detail Panel (on Row Selection)

The detail panel provides four tabs:

- **Properties**: Key-value display of all properties for the selected object
- **Links**: Tables of linked objects for each link type, organized by
  relationship
- **Actions**: Auto-generated forms for available actions from the bundle
- **Graph**: Network visualization of the object's neighbourhood (requires
  vertexR + visNetwork)

## Link Traversal: The Key Feature

Link traversal is what makes this an *ontology* explorer, not just a
database browser. The left sidebar's Link Panel shows all link types
where the current object type is either the `from` or `to` side.

For example, with the aviation demo:

- Viewing **Airport**: you see incoming links from FlightRoute
  (RouteOrigin, RouteDestination)
- Clicking **"<- Search Around"** on RouteOrigin: switches to
  FlightRoute view, filtered to routes that originate at the selected
  airport
- From **FlightRoute**: you see outgoing links to Airport and Airline
- Clicking **"Traverse ->"** on RouteOperator: switches to Airline view,
  filtered to the airline operating the selected route

Users navigate relationships naturally, following links across object
types without writing any queries.

## Action Execution

Actions are executed using `actionTypesR`. The recommended approach is to
pass an `ActionContext`:

```{r action-context, eval=FALSE}
library(actionTypesR)

action_ctx <- action_context(b, con)
action_ctx <- register_handler(action_ctx, "UpdateAirportStatus",
  function(conn, action, params, targets) {
    DBI::dbExecute(conn,
      "UPDATE airports SET status = ? WHERE airport_id = ?",
      list(params$new_status, targets[[1]]))
    list(status = "success")
  }
)

explore_ontology(b, con, action_ctx = action_ctx)
```

The deprecated `action_handlers` argument is still supported for backward
compatibility but will emit a warning.

## Concept Columns (Optional)

When `conceptR` is available, you can pass a function to evaluate concepts
and display them as columns in the results table:

```{r concept-cols, eval=FALSE}
explore_ontology(b, con,
  concept_cols_fn = function(type_id) {
    if (type_id == "Airport") {
      # Return a data frame with .pk and boolean concept columns
      data.frame(
        .pk = c("JFK", "LAX", "LHR"),
        busy_airport = c(TRUE, TRUE, FALSE)
      )
    } else {
      NULL
    }
  }
)
```

## Graph Neighbourhood (Optional vertexR)

If `vertexR` and `visNetwork` are installed, the Graph tab shows an
interactive network visualization of the selected object's neighbourhood
(depth = 2). This uses `vx_neighbors()` and `vx_plot()` from vertexR.

The graph panel only appears when:
- `vertexR` is installed
- `visNetwork` is installed
- The bundle has at least one link type defined

## Dark Mode

Click the moon/sun icon in the top-right corner to toggle between light
and dark themes. Requires Bootstrap 5 (provided by `bslib`).

## Module Architecture

The application is built as composable Shiny modules with a clear
reactive data flow:

```
type_selector -> selected_object_type_id (reactive)
    |
    v
filter_panel -> active_filters (reactive list)
    |
    v
objectSetsR::os_filter() -> filtered ObjectSet
    |
    v
objectSetsR::os_collect() -> filtered_data (reactive df)
    |
    v
results_table -> selected_row (reactive)
    |
    v
detail_panel -> properties / links / actions / graph tabs
    |
    v
link_panel -> traverse action -> updates type_selector + filters
```

Each module is self-contained with its own UI and server functions:

| Module               | Purpose                              |
|---------------------|--------------------------------------|
| `mod_type_selector`  | Object type dropdown + search        |
| `mod_filter_panel`   | Auto-generated property filters      |
| `mod_results_table`  | Interactive DT results table         |
| `mod_detail_panel`   | Properties/links/actions/graph tabs  |
| `mod_link_panel`     | Link traversal controls              |
| `mod_action_form`    | Action parameter form + submit       |

## Data Access via objectSetsR

All data access goes through `objectSetsR`. The `os_helpers.R` module provides
bridging functions that translate UI filter specs into `os_filter()` calls:

```{r os-helpers, eval=FALSE}
# Filter specifications from the UI are applied like this:
os <- object_set(ctx, "Airport")
os <- apply_filter_to_os(os, list(
  property_id = "country",
  type = "select_multi",
  value = c("United States", "United Kingdom")
))
result <- os_collect(os)

# Link traversal uses os_traverse() and os_search_around()
linked <- traverse_link_os(ctx, link_type, source_ids, "forward")
```

## Dependencies

**Required:**

- **ontologySpecR**: Ontology type definitions and bundle I/O
- **objectSetsR**: Lazy query layer (all data access)
- **actionTypesR**: Action execution
- **DBI**: Database connectivity
- **shiny**: Web application framework
- **bslib**: Bootstrap 5 theming (dark mode)
- **DT**: Interactive tables
- **dplyr**, **rlang**, **jsonlite**: Data manipulation utilities

**Optional:**

- **vertexR** + **visNetwork**: Graph neighbourhood visualization
- **conceptR**: Concept evaluation columns
