---
title: "objectExplorerR: Interactive Ontology Object Explorer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{objectExplorerR: Interactive Ontology Object Explorer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`objectExplorerR` is the first interactive user-facing application in the
ontology R ecosystem. It provides a Shiny-based exploratory UI that lets
users search, filter, traverse links, and inspect ontology objects
visually — without writing code.

This corresponds to Palantir's Object Explorer concept: visual query
composition and exploration views with relationship navigation.

## Quick Start

```{r launch-example, eval=FALSE}
library(ontologySpecR)
library(objectExplorerR)
library(DBI)

# Load an ontology bundle
b <- read_bundle("path/to/your-bundle.json")

# Connect to your database
con <- dbConnect(RSQLite::SQLite(), "your-database.db")

# Launch the explorer
explore_ontology(b, con)
```

## Setup with the Aviation Demo

The ontologySpecR package ships with an aviation demo bundle:

```{r setup-demo, eval=TRUE}
library(ontologySpecR)

# Read the aviation demo bundle
b <- read_bundle(system.file("examples", "aviation-demo.json",
                              package = "ontologySpecR"))

# The bundle defines 3 object types
cat("Object types:", paste(vapply(b$objects, function(o) o$id, ""), collapse = ", "), "\n")

# And 3 link types
cat("Link types:", paste(vapply(b$links, function(l) l$id, ""), collapse = ", "), "\n")

# And 1 action type
cat("Action types:", paste(vapply(b$actions, function(a) a$id, ""), collapse = ", "), "\n")
```

## Application Layout

The explorer is organized into four main panels:

```
+------------------------------------------------------+
|  [Object Type Selector]  [Search: ________]  [Refresh]|
+------------------------+-----------------------------+
|                        |                             |
|   FILTER PANEL         |    RESULTS TABLE            |
|                        |                             |
|  Property filters      |    Interactive DT table     |
|  (auto-generated       |    with sort, page,         |
|   from object type     |    select rows              |
|   properties)          |                             |
|                        +-----------------------------+
|  [Apply] [Clear]       |                             |
|                        |    DETAIL PANEL              |
|   LINK PANEL           |    (shown when row selected)|
|                        |                             |
|  Available links       |    - Properties tab         |
|  [Traverse ->]         |    - Links tab              |
|  [<- Search Around]    |    - Actions tab            |
|                        |    - Graph tab              |
+------------------------+-----------------------------+
```

### Top Bar: Object Type Selector + Search

The top bar contains a dropdown populated from the bundle's object types
(using display names), a free-text search box that filters across all
string properties, and a refresh button.

### Filter Panel (Left Sidebar)

Filters are **auto-generated** from the selected object type's property
definitions:

| Property Type | Widget                 |
|--------------|------------------------|
| `string`     | Text input or dropdown |
| `integer`    | Range slider           |
| `number`     | Range slider           |
| `boolean`    | Checkbox               |
| `date`       | Date range picker      |
| `datetime`   | Date range picker      |
| `json`       | (not filterable)       |

For string properties with few distinct values (< 20), the filter
automatically uses a dropdown instead of a text input.

Numeric sliders are initialized with the actual min/max from the data.

### Results Table (Main Area)

An interactive DataTable showing filtered objects with sortable columns,
pagination, and row selection. Clicking a row opens the detail panel.

### Detail Panel (on Row Selection)

The detail panel provides four tabs:

- **Properties**: Key-value display of all properties for the selected object
- **Links**: Tables of linked objects for each link type, organized by
  relationship
- **Actions**: Auto-generated forms for available actions from the bundle
- **Graph**: Network visualization of the object's neighborhood (requires
  vertexR + visNetwork)

## Link Traversal: The Key Feature

Link traversal is what makes this an *ontology* explorer, not just a
database browser. The left sidebar's Link Panel shows all link types
where the current object type is either the `from` or `to` side.

For example, with the aviation demo:

- Viewing **Airport**: you see incoming links from FlightRoute
  (RouteOrigin, RouteDestination)
- Clicking **"<- Search Around"** on RouteOrigin: switches to
  FlightRoute view, filtered to routes that originate at the selected
  airport
- From **FlightRoute**: you see outgoing links to Airport and Airline
- Clicking **"Traverse ->"** on RouteOperator: switches to Airline view,
  filtered to the airline operating the selected route

Users navigate relationships naturally, following links across object
types without writing any queries.

## Optional Integrations

### actionTypesR

If `actionTypesR` is installed, the Actions tab uses its full execution
engine (`submit_action()`). Without it, actions fall back to directly
calling handler functions you provide:

```{r action-handlers, eval=FALSE}
explore_ontology(b, con, action_handlers = list(
  UpdateAirportStatus = function(conn, action, params, targets) {
    DBI::dbExecute(conn,
      "UPDATE airports SET status = ? WHERE airport_id = ?",
      list(params$new_status, targets[[1]]))
  }
))
```

### vertexR

If `vertexR` and `visNetwork` are installed, the Graph tab shows an
interactive network visualization of the selected object's neighborhood
(depth = 1). This uses `vx_neighbors()` and `vx_plot()` from vertexR.

## Module Architecture

The application is built as composable Shiny modules with a clear
reactive data flow:

```
type_selector -> selected_object_type_id (reactive)
    |
    v
filter_panel -> active_filters (reactive list)
    |
    v
query_helpers::fetch_objects() -> filtered_data (reactive df)
    |
    v
results_table -> selected_row (reactive)
    |
    v
detail_panel -> properties / links / actions / graph tabs
    |
    v
link_panel -> traverse action -> updates type_selector + filters
```

Each module is self-contained with its own UI and server functions:

| Module               | Purpose                              |
|---------------------|--------------------------------------|
| `mod_type_selector`  | Object type dropdown + search        |
| `mod_filter_panel`   | Auto-generated property filters      |
| `mod_results_table`  | Interactive DT results table         |
| `mod_detail_panel`   | Properties/links/actions/graph tabs  |
| `mod_link_panel`     | Link traversal controls              |
| `mod_action_form`    | Action parameter form + submit       |

## Query Helpers

The package includes a `query_helpers.R` module that builds parameterized
SQL directly from ontologySpecR metadata. This works with any DBI-compatible
database without requiring objectSetsR:

```{r query-helpers, eval=FALSE}
# Fetch objects with filters
data <- fetch_objects(con, airport_type,
  filters = list(
    list(property_id = "country", type = "text", value = "United States")
  ),
  search_text = "International"
)

# Fetch linked objects
linked <- fetch_linked(con, bundle, route_origin_link,
                        source_object_ids = c("R001"),
                        direction = "forward")

# Get filter widget data
countries <- fetch_distinct_values(con, airport_type, "country")
elev_range <- fetch_range(con, airport_type, "elevation_ft")
```

## Dependencies

The package requires only three core dependencies beyond base R:

- **ontologySpecR**: Ontology type definitions and bundle I/O
- **DBI**: Database connectivity
- **shiny**: Web application framework
- **DT**: Interactive tables
- **dplyr**, **rlang**, **jsonlite**: Data manipulation utilities

Optional packages (`actionTypesR`, `vertexR`, `visNetwork`, `bslib`)
add features but are not required. The explorer works standalone with
just the core dependencies.
